---
title: "codigos_utiles"
author: "Nicolas"
date: "2025-07-29"
output: html_document
---

-----------------------------------------------------------
In this future PDF, i´ll going to develop my personal book
that is going to help me, to become in the best r person, r
man or something like that, i am going to write all the code 
in English just for practice.
# (CTRL+ ALT + I) to create a markdown cell,

para esteban puedes guiarte de aca hay mucho que falta, pero
es un buen inicio
-----------------------------------------------------------
```{r}
# If you use the function without parameters, show you the real math function behind the word
sd # this is one example
```
# The first section is very usefull codes (regular or basic)

```{r datos}
# Standard deviation
print("stanrd deviation of children")
sd(datos$Hijos) # only numeric vectors
print("--------------------------")
# mean
print("schooling mean")# mean the classic meaning
mean(datos$Escolaridad)
print("--------------------------")
#median
print("children median")
median(datos$Hijos) # median of data, that means the half of data, where is the midpoint
print("--------------------------")
#variance
print("variance in schooling")
var(datos$Escolaridad) # variance of schooling, is the dispersion of data against the average.
print("--------------------------")
# quantile
print("Quantile in children")
quantile(datos$Hijos)# split the data in 5 different sections, called quantile
print("--------------------------")
# unique 
print("unique like group by")
unique(datos$EstadoCivil) # that is useful when someone ask for how many in a group
print("--------------------------")
```
# this second section is for test and models (intermidian)
```{r datos}
# models
print("models of series(series it means to a different models)")
lm# this is linear regression
glm # this is general form of lm
aov # this is anova a comparative of average from 2 or more different samples. with summary is very useful here because, it have a lot of results.
print("--------------------------")
# test
chisq.test(datos$Hijos, datos$Escolaridad) # this test is for chi square distribution, relation between two variables and perfect distribution bound fitting
t.test(datos$Hijos, datos$Escolaridad) # it use for compare average (mu) for expected results
# parametric test
shapiro.test()
ks.test()
# non parametric test
wilcox.test()
# correlation
cor.test(x, y, method="pearson") # in method can be use " pearson, Kendall, spearman"
```
# in this third section is for plots all tipe of plots and all type of settings that we can use.
```{r datos}
# plot
x <- c(datos$Hijos)
y <- c(datos$Escolaridad)
plot(x, y, col= "orange", type= "l", lty = 5, frame.plot = FALSE) 
grid(nx = NA, ny = NULL, col = "#FDF5E6", lty = 3, lwd = 1.5) # grid for the graphic just put at the end
# for plot just x (horizontal), y (vertical)
```
-----------------------------------------------------------------------------
                      SETINGS FOR PLOT  plot()
-----------------------------------------------------------------------------
Type: type of union of the graphic 
"p": points
"l": lines
"b": both, lines and points
example: type = "p"
-----------------------------------------------------------------------------
main : title of the plot
main = ""
example: main "Esteban is fool"
-----------------------------------------------------------------------------
col: color of the graphic
col = "red", can be a specific color
col = rgb(0,0,0) can be rgb
col = #FF5733 can be rgb in hexadecimal
col = palette(1,2,3) can be palette in r
col = c("red", "blue", etc) can be combine in multiple colors
example col = "orange"
-----------------------------------------------------------------------------
pch: point symbol
depends of the number can be a different symbol
can´t use bg
0	Square	No	
1	Circle	No	
2	Triangle up	No	
3	Plus sign	No
4	Cross (×)	No	
5	Diamond	No	
6	Triangle down	No
can use background--- 21- 25 col affects the border, bg the center
15	Filled 	
16	Filled 
17	Filled 
18	Filled 
21–25	Point + Border	
example: pch = 0
-----------------------------------------------------------------------------
lty: line type
0	Blank	Invisible line (no rendering)
1	Solid	Default solid line
2	Dashed	Elegant separation or grouping
3	Dot ted	Light touch, subtle boundary
4	Dot dash	Visual rhythm, noticeable
5	Long dash	Slower pacing in visuals
6	Two dash	Minimalist dual-dash
example: lty = 4
-----------------------------------------------------------------------------
lwd: thickness or width of the line
a scale
example: lwd = 1.5
-----------------------------------------------------------------------------
bg: background color of the symbol no the back
it the same structure that in colors
-----------------------------------------------------------------------------
cex: scale of point
example: cex = 1.5
-----------------------------------------------------------------------------
axes: show axes no or yes
axes = TRUE or FALSE
example: axes = FALSE
-----------------------------------------------------------------------------
frame plot: a box around the axes
False or true
example: frame.plot = TRUE
-----------------------------------------------------------------------------
```{r}
# histogram or bar plot (most useful bar plot because put labels to the categorization)

barplot(table(datos$Genero),
        col = c("blue", "pink"),   
        border = "white",
        main = "Gender Distribution",
        xlab = "Gender",
        ylab = "Count", names.arg = c("Female", "Male"))
#grid(col = "black")
# for plot just x (horizontal), y (vertical)

```
-----------------------------------------------------------------------------
                SETTINGS FOR HISTOGRAM hist()
-----------------------------------------------------------------------------
breaks: quantity of splits
can be specific like
breaks = c(0, 5, 10, 15, 20)
or sequential 
example: breaks = seq(0,100 (start and end), by = 10 (that means the frequency))
-----------------------------------------------------------------------------
main, xlab, ylab: labels to write
-----------------------------------------------------------------------------
xlim, ylim: specific limits for the scale in the axis
xlim = c(0, 100) start and end
example: ylim = c(-8, -9)
-----------------------------------------------------------------------------
col: color of the graphic
col = "red", can be a specific color
col = rgb(0,0,0) can be rgb
col = #FF5733 can be rgb in hexadecimal
col = palette(1,2,3) can be palette in r
col = c("red", "blue", etc) can be combine in multiple colors
example col = "orange"
-----------------------------------------------------------------------------
border: border color
the same that col
-----------------------------------------------------------------------------
freq: in theory count the population on the groups
example: freq = TRUE
-----------------------------------------------------------------------------
labels: you can change the names of the labels for category
example: names.arg = c("Female", "Male")
-----------------------------------------------------------------------------
-----------------------------------------------------------------------------
```{r}
# boxplot
boxplot(datos$Escolaridad ~ datos$Genero, names = c("Female", "Male"), col = c("pink", "blue"), border = "gray", width= c(0.6, 1.8), notch = TRUE)
```
-----------------------------------------------------------------------------
                          SETTINGS FOR BOXPLOT
-----------------------------------------------------------------------------
main, xlab, ylab: labels to write
labels basic
example: main = "project", xlab = "x"
-----------------------------------------------------------------------------
names
you can change the name of the label in each category
example: names = c("Female", "Male")
-----------------------------------------------------------------------------
horizontal
flip the content to horizontal
example: horizontal = TRUE
-----------------------------------------------------------------------------
col: color of the graphic
col = "red", can be a specific color
col = rgb(0,0,0) can be rgb
col = #FF5733 can be rgb in hexadecimal
col = palette(1,2,3) can be palette in r
col = c("red", "blue", etc) can be combine in multiple colors
example col = "orange"
-----------------------------------------------------------------------------
notch: split the ox in their average it looks like a polygon
example: notch = true
-----------------------------------------------------------------------------
border
the same of col
-----------------------------------------------------------------------------
freq: 
in theory count
example: freq = TRUE
-----------------------------------------------------------------------------
width: the size of the box
example width = c(2.5 (first box), 2.5 (second box))
-----------------------------------------------------------------------------
-----------------------------------------------------------------------------
```{r}
# Plot Density
plot(density(datos$Escolaridad))
```
-----------------------------------------------------------------------------
                SETTINGS FOR HISTOGRAM DENSITY
-----------------------------------------------------------------------------
main, xlab, ylab: labels to write
example: main = "project", xlab = "x"
-----------------------------------------------------------------------------
xlim, ylim: specific limits for the scale in the axis
xlim = c(0, 100) start and end
example; ylim = c(-8, -9)
-----------------------------------------------------------------------------
col: color of the graphic
col = "red", can be a specific color
col = rgb(0,0,0) can be rgb
col = #FF5733 can be rgb in hexadecimal
col = palette(1,2,3) can be palette in r
col = c("red", "blue", etc) can be combine in multiple colors
example col = "orange"
-----------------------------------------------------------------------------
lwd: thickness or width of the line
a scale
example: lwd = 1.5
-----------------------------------------------------------------------------
lty: line type
0	Blank	Invisible line (no rendering)
1	Solid	Default solid line
2	Dashed	Elegant separation or grouping
3	Dot ted	Light touch, subtle boundary
4	Dot dash	Visual rhythm, noticeable
5	Long dash	Slower pacing in visuals
6	Two dash	Minimalist dual-dash
example: lty = 4
-----------------------------------------------------------------------------
-----------------------------------------------------------------------------
```{r}

system("git --version")
```

















